---
title: "sisbidDay3"
author: "Angie Boysen"
date: "July 13, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##dplyr

```{r}
install.packages('gapminder')
library(dplyr)
library(gapminder)
str(gapminder)
gtbl = gapminder

```

filter
```{r}
glimpse(gtbl)
filter(gtbl, lifeExp < 29)
filter(gtbl, country == "Rwanda")
```


look at columns with select
```{r select}
select(gtbl,country,pop,continent)

```

look at arrange for re-ordering rows
```{r}
# arrange(gtbl,pop)
# arrange(gtbl,desc(pop)) #Order in reverse
arrange(gtbl, year, lifeExp) #Order by two things
```

Creating new variables with mutate
```{r}
gtbl = mutate(gtbl, newVar = (lifeExp / gdpPercap), newVar2 = newVar*2)
select(gtbl,lifeExp,gdpPercap,newVar)
distinct(gtbl) # spits out all unique rows - can detect duplucate rows/errors in database
```

sampling data with dplyr
```{r}
# set.seed(12345) #makes the randomness the same every time you run this
sample_n(gtbl,3) # 3 random rows
sample_frac(gtbl,0.5) # random 50%
```

## piping

all of these dplyr functions have the first argument as the dataframe
So if you chain a bunch of these together it works nicely because the output is always a dataframe that you can input into another function.

%>% -- read as "then"

These two pieces of code do the same thing:
```{r}
head(gtbl)
gtbl %>% head()
```

Example: show me a random sample of the data for asian countries with life expectancy < 65.
```{r}
gtbl1 <- gtbl[gtbl$continent =="Asia",]
gtbl2 <- gtbl1[gtbl1$lifeExp < 65,]
gtbl3 <- gtbl2[sample(1:dim(gtbl2)[1],size=10),]
gtbl3

## OR
random_sample <- gtbl %>% filter(continent =="Asia") %>%
     filter(lifeExp <65) %>%
     sample_n(10)
random_sample

## Can do Asia or Africa

random_sample <- gtbl %>% filter(continent %in% c("Asia","Africa")) %>%
     filter(lifeExp <65) %>%
     sample_n(10)
random_sample
```

Example2: Split, apply, combine

What is the average life expectancy by continent?

```{r}
gtbl %>% group_by(continent) %>% summarize(aveLife = mean(lifeExp))

## can do more than one thing
gtbl %>% group_by(continent) %>% 
     summarize(aveLife = mean(lifeExp), sdLife = sd(lifeExp))

## can definte your own fucntion to use
mean2 = function(x){mean(x/2)}

```

